A flexible integer linear programming formulation for scheduling clinician on-call service in hospitals

### 2019-09-16 paper/paper.methods.tex
Fix methods section

In this section, we present an application of integer linear programming to solve the clinician
% JK: there are a few variations of
%     "nurse scheduling problem", "scheduling problem" "clinician scheduling problem"
%     floating around. Can you do a consistency check throughout?
scheduling problem presented in Section~\ref{sec:problem}. First, we describe
% JK: Also, while the intro gives a reference for ILP,
%     I think you should define it here. e.g. It is not clear to the naive reader (me!)
%     that ILP needs an objective function
%     or (most importantly) how ILP works to satisfy multiple constraints.
the sets, indices and variables necessary for the formulation of the problem. We
then write the constraints given in Table~\ref{tbl:constraint-summary} as
mathematical equations, and define the objective function of the linear program.

\subsection{Sets and Indices}\label{sec:meth-sets-indices}
We denote the set of all services %or I think services is good (vs.
%services/divisions) as you define two types of services (ID vs. HIV) in the
%problem statement
% JK: same thing, consistency check with division / service
%     I think maybe you already did this...
that clinicians in a single department can provide as $\mathcal{S}$. The
following formulation assumes that all clinicians are able to provide all
% JK: but could you not also just set some clinician-service max-mins zero
%     to deal with if some clinicians are not able to provide some services?
services. The set of all clinicians in the department is denoted as
$\mathcal{C}$. The sets of blocks and weekends
% JK: removed "will be assigned to" since on first read I confused it with
%     the binary assignment variable X below.
are denoted as $\mathcal{B}$ and $\mathcal{W}$ respectively. The block size
used in our experiments is 2 weeks, but the following LP formulation does not
require a particular size for blocks, and so it can be adapted for other cases.
%The size of a block is not constrained and can be adapted %how?
%to the needs of the given department. 
A subset of weekends are denoted $\mathcal{L}$, corresponding to established
long/holiday weekends such as the Canadian Civic Day weekend in August. Lastly,
% JK: I'm being picky, but maybe a non-Canadian holiday like Thanksgiving as the e.g.
the block and weekend request-off dates of clinicians as are denoted as the following
subsets of all blocks and weekends: $\mathcal{BR}_c$ and $\mathcal{WR}_c$, respectively.
% JK: keeping "x is denoted as x" sentence structure
% JK: Also, I'm generally not a fan of multiple-letter variables since they can be 
%     easily confused for multiplication. Can you find another notation?
For instance, if clinician $c$'s request-off dates intersect with blocks 1 and 2, and weekend
1, then $\mathcal{BR}_c = \{1, 2\}$ and $\mathcal{WR}_c = \{1\}$. Table~\ref{tbl:sets-indices} presents a summary of the sets and indices described.

\begin{table}[h]
	\centering
  \caption{Description of sets and indices in the problem}%
  \label{tbl:sets-indices}
	\begin{tabular}{ c c l }
		\toprule
		\textbf{Set}                         & \textbf{Index} & \textbf{Description}  
		\\ \midrule
		$\mathcal{S} = \{1, \ldots, S \}$    & $s$            & services              
		\\
		$\mathcal{C} = \{1, \ldots, C \}$    & $c$            & clinicians            
		\\
		$\mathcal{B} = \{1, \ldots, B \}$    & $b$            & blocks                
		\\
		$\mathcal{W} = \{1, \ldots, W \}$    & $w$            & weekends              
		\\
		$\mathcal{L} \subset \mathcal{W}$    &                & long weekends         
		\\
		$\mathcal{BR}_c \subset \mathcal{B}$ &                & block requests of
		clinician $c$   \\
		$\mathcal{WR}_c \subset \mathcal{W}$ &                & weekend requests of
		clinician $c$ \\
    \bottomrule
	\end{tabular}

@@

\subsection{Objectives}\label{sec:meth-objectives}
As described in Section~\ref{sec:problem}, the soft constraints of the clinician
scheduling problem include: satisfying clinician block off requests (BR),
satisfying clinician weekend off requests (WR), and assigning weekends closer to
blocks (BWA). We convert these soft constraints into linear objective functions
of the binary variables defined in Section~\ref{sec:meth-variables}. Objectives
(\ref{eqn:obj-block-requests}) and (\ref{eqn:obj-weekend-requests}) can be
written as the following linear functions:  %minor suggestion - consider
%numbering the objectives for ease later when saying things like "these two
%objectives"...or writing the full name of the objectives instead of 'this' and
%'these' as sometimes can get confusing what the 'this' and the 'these' refer
%to. 
\begin{align}
&Q_1(X) = \sum_{c=1}^{C} \sum_{b=1}^{B} \sum_{s=1}^{S}
{(-1)}^{\ind(b\,\in\,\mathcal{BR}_c)}\cdot X_{c, b, s} \tag{BR}
\label{eqn:obj-block-requests}\\
&Q_2(Y) = \sum_{c=1}^{C} \sum_{w=1}^{W}
{(-1)}^{\ind(w\,\in\,\mathcal{WR}_c)}\cdot Y_{c, w} \tag{WR}
\label{eqn:obj-weekend-requests}
\end{align}
where $\ind(P)$ is the indicator function that has value 1 when predicate $P$
holds and 0 otherwise. In the above two objectives, we penalize any assignments
that conflict with a block or weekend request, and aim to maximize the
non-conflicting assignments.

@@

In order to optimize all objectives simultaneously, we optimize a weighted sum
of the normalized objective functions,
% JK: This only considers the soft constraints.
%     Should this maximization not also be subject to the Constraints in 3.3?
\begin{equation}
\max_{X, Y, Z} \alpha \bar{Q}_1(X) + \beta \bar{Q}_2(Y) + (1 - \alpha - \beta)
\bar{Q}_3(Z)
\end{equation}
% JK: I would use $\alpha_1, \alpha_2, \alpha_3$ here which could take any value,
%     and then not necessary to have any bounds like \sum_i^3 \alpha_i = 1,
%     though you could enforce it to compare across permutations of \alpha
where $\bar{Q}_i$ is the normalization of objective $Q_i$, and $0 \leq \alpha,
\beta \leq 1$. This method guarantees an optimal solution to be Pareto optimal~\cite{stanimirovic_linear_2011}.
% JK: How can ILP be used solve such problems? Can you briefly describe?

>>>>

In this section, we present an application of integer linear programming to solve the clinician
% JK: there are a few variations of
%     "nurse scheduling problem", "scheduling problem" "clinician scheduling problem"
%     floating around. Can you do a consistency check throughout?
scheduling problem presented in Section~\ref{sec:problem}. 
An integer linear program (ILP) consists of a linear objective function and linear constraints,
with integer-valued variables. The optimal solution to the ILP must lie within the space
defined by the constraints while also maximizing the objective value.
First, we describe
% JK: Also, while the intro gives a reference for ILP,
%     I think you should define it here. e.g. It is not clear to the naive reader (me!)
%     that ILP needs an objective function
%     or (most importantly) how ILP works to satisfy multiple constraints.
the sets, indices and variables necessary for the formulation of the problem. We
then write the constraints given in Table~\ref{tbl:constraint-summary} as
linear functions of the variables, and define the objective function of the ILP.

\subsection{Sets and Indices}\label{sec:meth-sets-indices}
We denote the set of all services %or I think services is good (vs.
%services/divisions) as you define two types of services (ID vs. HIV) in the
%problem statement
% JK: same thing, consistency check with division / service
%     I think maybe you already did this...
that clinicians in a single department can provide as $\mathcal{S}$. 
%The following formulation assumes that all clinicians are able to provide all services. 
% JK: but could you not also just set some clinician-service max-mins zero
%     to deal with if some clinicians are not able to provide some services?
The set of all clinicians in the department is denoted as
$\mathcal{C}$. The sets of blocks and weekends
% JK: removed "will be assigned to" since on first read I confused it with
%     the binary assignment variable X below.
are denoted as $\mathcal{B}$ and $\mathcal{W}$ respectively. The block size
used in our experiments is 2 weeks, but the following LP formulation does not
require a particular size for blocks, and so it can be adapted for other cases.
%The size of a block is not constrained and can be adapted %how?
%to the needs of the given department. 
A subset of weekends are denoted $\mathcal{L} \subset \mathcal{W}$, corresponding to established
long/holiday weekends such as the Thanksgiving weekend. Lastly,
% JK: I'm being picky, but maybe a non-Canadian holiday like Thanksgiving as the e.g.
the block and weekend time-off requests of clinicians are denoted as the following
subsets of all blocks and weekends: $\mathcal{U}_c \subset \mathcal{B}$ and $\mathcal{V}_c \subset \mathcal{W}$, respectively.
% JK: keeping "x is denoted as x" sentence structure
% JK: Also, I'm generally not a fan of multiple-letter variables since they can be 
%     easily confused for multiplication. Can you find another notation?
For instance, if clinician $c$'s time-off requests intersect with blocks 1 and 2, and weekend
1, then $\mathcal{U}_c = \{1, 2\}$ and $\mathcal{V}_c = \{1\}$. Table~\ref{tbl:sets-indices} presents a summary of the sets and indices described.

\begin{table}[h]
	\centering
  \caption{Description of sets and indices in the problem}%
  \label{tbl:sets-indices}
	\begin{tabular}{ c c l }
		\toprule
		\textbf{Set}                         & \textbf{Index} & \textbf{Description}  
		\\ \midrule
		$\mathcal{S} = \{1, \ldots, S \}$    & $s$            & services              
		\\
		$\mathcal{C} = \{1, \ldots, C \}$    & $c$            & clinicians            
		\\
		$\mathcal{B} = \{1, \ldots, B \}$    & $b$            & blocks                
		\\
		$\mathcal{W} = \{1, \ldots, W \}$    & $w$            & weekends              
		\\
		$\mathcal{L} \subset \mathcal{W}$    &                & long weekends         
		\\
		$\mathcal{U}_c \subset \mathcal{B}$ &                & block requests of
		clinician $c$   \\
		$\mathcal{V}_c \subset \mathcal{W}$ &                & weekend requests of
		clinician $c$ \\
    \bottomrule
	\end{tabular}

@@

\subsection{Objectives}\label{sec:meth-objectives}
As described in Section~\ref{sec:problem}, the soft constraints of the clinician
scheduling problem include: satisfying clinician block off requests (BR),
satisfying clinician weekend off requests (WR), and assigning weekends closer to
blocks (BWA). We convert these soft constraints into linear objective functions
of the binary variables defined in Section~\ref{sec:meth-variables}. Objectives
(\ref{eqn:obj-block-requests}) and (\ref{eqn:obj-weekend-requests}) can be
written as the following linear functions:  %minor suggestion - consider
%numbering the objectives for ease later when saying things like "these two
%objectives"...or writing the full name of the objectives instead of 'this' and
%'these' as sometimes can get confusing what the 'this' and the 'these' refer
%to. 
\begin{align}
&Q_1(X) = \sum_{c=1}^{C} \sum_{b=1}^{B} \sum_{s=1}^{S}
{(-1)}^{\ind(b\,\in\,\mathcal{U}_c)}\cdot X_{c, b, s} \tag{BR}
\label{eqn:obj-block-requests}\\
&Q_2(Y) = \sum_{c=1}^{C} \sum_{w=1}^{W}
{(-1)}^{\ind(w\,\in\,\mathcal{V}_c)}\cdot Y_{c, w} \tag{WR}
\label{eqn:obj-weekend-requests}
\end{align}
where $\ind(P)$ is the indicator function that has value 1 when predicate $P$
holds and 0 otherwise. In the above two objectives, we penalize any assignments
that conflict with a block or weekend request, and aim to maximize the
non-conflicting assignments.

@@

In order to optimize all objectives simultaneously, we optimize a weighted sum
of the normalized objective functions,
% JK: This only considers the soft constraints.
%     Should this maximization not also be subject to the Constraints in 3.3?
\begin{equation}
\max_{X, Y, Z} \alpha_1 \bar{Q}_1(X) + \alpha_2 \bar{Q}_2(Y) + \alpha_3
\bar{Q}_3(Z)
\end{equation}
% JK: I would use $\alpha_1, \alpha_2, \alpha_3$ here which could take any value,
%     and then not necessary to have any bounds like \sum_i^3 \alpha_i = 1,
%     though you could enforce it to compare across permutations of \alpha
subject to the constraints defined in Section \ref{sec:meth-constraints},
where $\bar{Q}_i$ is the normalization of objective $Q_i$, and $0 \leq \alpha_i \leq 1$. 
This method guarantees an optimal solution to be Pareto optimal~\cite{stanimirovic_linear_2011}.

Currently, the most efficient approach to finding an exact solution for 
an ILP is called Branch-and-Cut. This method involves iteratively solving 
relaxed versions of the ILP, then constraining the relaxed problems and
considering various sub-problems until it finds integral solutions to the 
original ILP~\cite{mitchell_branch-and-cut_2002}.
The intermediate relaxations allow the variables to take on real values, 
and can be solved efficiently using the Simplex method~\cite{shamir_efficiency_1987}. The 
complexity of finding an optimal integral solution thus lies in the 
branching search structure of the approach.

% JK: How can ILP be used solve such problems? Can you briefly describe?
